Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-06-29 POPL/paper.tex
Abstract + introduction condensing

% -*- fill-column: 80; -*-
\documentclass[sigconf,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}

\usepackage{todonotes}
\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

\title{Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}
\author{Steven Schaefer}
\affiliation{
  \department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\section{A Syntax-Independent Notion of Syntax}
\todoin{Rework intro from LICS draft with a focus on implementation}

\section{Kleene Categories}
\todoin{Tell a story winding through the use of KAs, action algebras, etc to
  motivate why we'd want categorical semantics for KCats}
\todoin{Incoporate Kleene Category definition}
\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}

\section{Automata}

\end{document}

>>>>

% -*- fill-column: 80; -*-
\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

\usepackage{pdfpages}

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\nat}{\mathbb{N}}

@@

\newif\ifdraft
\drafttrue
\newcommand{\steven}[1]{\ifdraft{\color{orange}[{\bf Steven says}: #1]}\fi}
\renewcommand{\max}[1]{\ifdraft{\color{blue}[{\bf Max says}: #1]}\fi}
\newcommand{\pedro}[1]{\ifdraft{\color{red}[{\bf Pedro says}: #1]}\fi}
\newcommand{\pipe}{\,|\,}

@@

\pagestyle{plain}

\pagebreak

\title{Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory}

\author{Steven Schaefer}
\affiliation{\department{Electrical Engineering and Computer Science}
  \institution{University of Michigan}
  \country{USA}
}
\email{stschaef@umich.edu}

@@

\setcounter{page}{1}

\section{Introduction}

\section{A Syntax-Independent Notion of Syntax}
The notion of formal language is central to the theory of parsing. A
\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
\emph{subset} of strings $L \subseteq \String$. This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.

That is, a recognizer should be extensionally equivalent to the indicator function
$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
that it is in the language $w \in L$. However, language recognition is
insufficient for specifying a parser --- a function that decomposes a string
into semantic components that adhere to the structure of a formal grammar. In
practice, parsers do more than simply return whether a string is valid or
invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
  trees are usually not materialized in memory, as semantic actions often perform
  computation over the trees while they are being constructed.}

Our first contribution is a novel and syntax-independent characterization of
formal grammar:

\begin{definition}
  \label{def:grammar}
  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
  $G : \String \to \Set$.
\end{definition}

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations are used to encode $\Set$. Our
  construction is polymorphic in choicce of a proper class of small sets, a
  Grothendieck universe, a type theoretic universe, or any similar foundation.} $G w$ of parse
trees showing that $w$ matches $G$. This definition of formal grammar serves as
a mathematical specification for a parser --- taking on the same role as
languages do for recognizers. In this sense, our new characterization of
grammars is just a proof-relevant generalization of the existing notion of
formal language.

By ranging over all input strings, notice that each grammar induces a language.
For each formal grammar $G$, define the language of accepted strings $L_{G}$ as,

\[ L_{G} = \{ w \in \String ~|~ G w~\text{is inhabited} \} \]

As they a simply sets, formal languages are naturally endowed with a partial
order via subset inclusion. Similarly, formal grammars are naturally endowed
with the structure of a \emph{category}. A morphism $f : G \to H$ between two
grammars is given by a family of functions

\[ f^{w} : G w \to H w \]

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer} : $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses --- precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

The above category of grammars can equivalently be described as the functor
category $\Set^{\String}$ --- where the set $\String$ is viewed as a discrete
category. That is, the objects in $\String$ are strings and the only morphisms
are identity morphisms. Such functor categories into $\Set$ --- often called
presheaf categories --- carry a remarkable amount of structure.

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal semantic
domain to reason about parsing. The slice 2-category over $\Set^{\String}$ is
then the appropriate setting to reason over notions of formal grammar.

\steven{Check details for 2-cat of grammar formalisms}

Our contributions are then:

\steven{These sound awkward but I'm painting with broad strokes rn}
\begin{itemize}
  \item A syntax-independent definition of formal grammar
  \item A dependent
    linear/non-linear (LNL) type theory to syntactically capture the semantics of parsing
  \item A categorification of Kleene Algebra
  \item Interpretations of the LNL type system for reasoning about formal grammars and formal languages
  \item The presentation of automata as grammars. By internalizing automata ---
        the operational view of a parser --- we may extract executable parsers
  \item A language implemented shallowly in Agda
\end{itemize}

\section{Kleene Category}
\steven{No changes pasted from LICS draft}
\steven{Are these now ``Chomsky categories''? or is that disinct?}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \caption{Kleene algebra axioms}
  \label{fig:axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we want to categorify the concept of
Kleene algebra and build on top of it in order to define an abstract
theory of parsing.  We start by defining \emph{Kleene categories}.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kleene categories.

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

This example illustrates how restricted Kleene algebras are in terms
of reasoning about languages. By only having at most one morphism
between objects, there is not a lot of information they can convey. In
this case, the only information you get is language containment. We
will see throughout this paper how the extra degrees of freedom
granted by having more morphisms give you more algebraic structure for
reasoning about languages.

For the next example, we see an unexpected connection with the theory
of substructural logics.

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics which allows hypotheses to
  be discarded or duplicated.
\end{example}

The proposed axioms are a direct translation of the Kleene algebra
axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family of initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
  uniqueness by showing that the diagram on top commutes if, and
  only if, the diagram on the bottom commutes:

% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
\[\begin{tikzcd}
	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
	\\
	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
	\\
	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
	\arrow["{[f,g]}", from=1-3, to=3-3]
	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
	\arrow["h"', from=6-1, to=6-3]
	\arrow["{[f', g']}", from=4-3, to=6-3]
	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
\end{tikzcd}\]

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{kozen1994}.

We are now ready to prove that our concept of formal grammars fits
nicely within our categorical framework. We start by presenting a
well-known construction from presheaf categories.

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z)
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}

\begin{lemma}[\cite{day1970}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}

  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts. Furthermore, the tensor
  is a left adjoint, i.e. it preserves colimits and, therefore, it is
  a distributive category.

  As for the Kleene star, since presheaf categories admit small colimits,
  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
  defined as the filtered colimit of the diagrams:

  From Theorem~\ref{th:kleeneclosed} it follows that these initial
  algebras satisfy the required isomorphisms and this concludes the
  proof.
\end{proof}

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

Much like in the posetal case, the abstract structure of a Kleene
category is expressive enough to synthetically reason about formal
languages. A significant difference between them is that while Kleene
algebras can reason about language containment, Kleene categories can
reason about parsing as well.

For the rest of the paper we will use the presheaf category
$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
guide when expanding abstract formalism presented in this section so
that it can handle more classes of languages beyond regular ones.

\paragraph{Beyond Simple Types}

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theory
point of view. This is limiting because type theories are tested
syntactic frameworks for manipulating complicated categories
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory.

\begin{definition}
  A model is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

\begin{theorem}
  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
\end{theorem}

\section{Related work}

\paragraph{Kleene Algebra}

Since the early works in the theory of formal languages, Kleene
algebras have played an important role in its development. They
generalize the operations known from regular languages by introducing
operations generalizing language composition, language union and the
Kleene star.  More generally, they are defined as inequational theory
where the inequality is meant to capture language containment. This
theory is extremely successful, having found applications in algebraic
path problems, theory of programming languages, compiler optimizations
and more.

A frequently fruitful research direction is exploring varying
extensions of Kleene algebras, Kleene algebra with tests (KAT) being
one of the most notable ones. Our approach is radically different from
most extensions, which usually aim at modifying or adding new
operations to Kleene algebras, but still keeping it as an inequational
theory. By adopting a category-theoretic treatment and allowing the
``order structure'' to encode more information than merely
inequalities, we were able to extend Kleene algebra to reason about
parsing as well.

\paragraph{TODO: more stuff}

%% That paper with Fritz Henglein about regexes as types

%% Vaughan Pratt on the residual operator

\section{Rest of paper}
\steven{Everything below still needs to be edited from my prelim paper}
\steven{Need to grab the canonicity proof from the LICS draft}

Substructural logics, like linear and separation logic, offer precise control
over how values from a type may be used, in particular how many times a value
may be used in a derivation. This control enables \emph{resource-sensitive}
reasoning, where we may treat a resource as \emph{consumed} after usage. In the
context of parsing, we may treat the characters composing a string as finite resources
that are consumed at parse-time.

Dependent type theory extends the simply-typed $\lambda$-calculus by allowing type
definitions to contain the occurrence of a term. In lieu of set-based
foundations, this extension can then be taken as a foundation for mathematics.
The expressive power unlocked by this insight has enabled the development of
dependently-typed proof assistants
such as Coq or Agda. Within each of these languages, systems may be given very
precise functional specifications of their behavior.

In 1995, Benton et al.
\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
that synthesized linear and non-linear logic. Later, Krishnaswami et al.
\cite{krishnaswami_integrating_2015} built upon this work to develop a type
system that unifies both linear and
dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
such settings, we have two sub-theories --- one linear and one non-linear. The
linear sub-theory embeds ordinary linear types without dependence, while the
non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
linear types with dependence lives at the boundary between these two fragments.
Benton et al.\ gives an accounting of how we may jump between these
fragments via a monoidal adjunction $F \dashv G$ between linear terms and
non-linear terms.

To the end of defining real verified
parsers, in our LNL theory we define a parser as a term,

\crefname{equation}{}{}
\begin{equation}
  \label{eq:parser}
  w : \String \vdash \mathsf{parse}_{g}(w) : g \oplus \top
\end{equation}

We will investigate the precise meaning of judgments like \cref{eq:parser} in
\cref{sec:tt}, but first let's get an intuitive feel for the meaning behind such a
term. On the left of the turnstile occurs a single string $w$, we may think of
this describing what resources are linearly available to us. Using each piece of
our the context on the left of the turnstile, we build a term
$\mathsf{parse}_{g}(w)$ of type $g \oplus \top$. Notably, the type being
instantiated is a sum type, and the data of which side we inject into denotes
whether the string is accepted or rejected by the parser. That is all to say, a
parser takes in an input string and either matches it to the desired grammar or
rejects the string.

Practical parsers are often implemented via an abstract automaton. For instance,
a parser may translate a grammar into a state machine recognizes the
same language as the original grammar, as the automata
point-of-view lends itself to straightforward execution. The
strength of our syntax is that it allows us to describe
these automata as grammars themselves. Moreover, the
equivalence between the grammar and automata presentations
is derivable as a proof term within the theory. This allows
us access to simultaneously internalization of the equivalence
between grammars and automata as well as the ability to
export to automaton implementations of a parser that are executable.

To realize an executable implementation of our type theory,
we shallowly embed it in Agda. Because our implementation is
defined as a function in Agda, we can leverage the Agda
ecosystem for code generation. Further, in lieu of
implementing a complete language from scratch, we make use of the
Agda proof system as a setting to guarantee correctness of
our generate parsers.

Within this paper, we give our progress in defining a verified
parser for regular expressions. The construction of such a parser entails
the definition of a parser for deterministic finite automata (DFA) grammars,
and an equivalence proof between DFAs and regular expressions. We have thus far
completed the construction of a parser for DFAs $D$,

\[
  w : \String \vdash \mathsf{parse}_{D}(w) : D \oplus \top
\]

and half of the equivalence proof between DFAs and regular expressions, both
formalized in Agda. Upon completion of the second half of this proof, we would
have a system where a user could simply give a regular grammar and then extract
an intrinsically verified parser for it. In future work, we
hope to show an analogous construction involving pushdown
automata (PDAs) and context-free grammars.

\section{A Syntax-Independent Notion of Syntax}
\label{sec:synindsyn}

% \max{this section doesn't flow naturally from the intro, which is all about the
%   type theory and parsing. This is where you want to outline in the intro and
%   possibly say here that we want to revisit ``formal grammar theory'' and then
%   we will make a type theory that has formal grammar theory as its semantics.}

In this section we will revisit some foundational ideas in formal
grammar theory, before giving a type
theory for which formal grammars serves as its semantics.

Formal languages and grammars are among some of the oldest and most studied areas in
theoretical computer science. The developments of the 1950s and 1960s gave rise
to expressive formalism for reasoning about grammars, and further practical
algorithms for constructing parsers for regular and context-free grammars.

A grammar $G$ is often described generatively as a set of production rules for
generating strings. Often this is presented as a finite set of nonterminal
symbols $N$, a finite set of terminal symbols $\Sigma$, and a finite set of rewriting
rules for substitution of nonterminals. In this setting, we continually apply
rewrite rules until we are left only with a string of terminals
$w \in \String$. We will give a more precise notion of
grammar in a moment, but for now let us take this intuitive view of grammars as
string generators.

One of the primary objects of study is that of a \emph{formal language} over
a fixed alphabet $\Sigma$. A language $L$ is a \emph{subset} of strings
$L \subseteq \String$. This notion of formal language gives an intuitive
set-based semantics for grammars. That is, we may interpret a grammar $G$ as the
language $L(G)$ that is generated by it. We are yet to give any precise
definition of what we mean by ``grammar'', although we note that any suitable
notion of grammar can be given a language semantics as above.

Formal languages serve as a basis for mathematically specifying a
\emph{recognizer} --- a
function $r : \String \to \Bool$ that determines if a string belongs to a
given language. However, they are insufficient for specifying a \emph{parser}
--- a function that decomposes a string into semantic components that adhere to
the structure of a formal grammar. In practice, parsers don't just tell you
if a string is valid or invalid. Instead, they emit structured data such as
abstract syntax trees (ASTs).

This motivates our first conceptual contribution, a simple and
syntax-independent notion of formal grammar:

\begin{definition}
  \label{def:grammar}
  A \emph{formal grammar} $G$ over a fixed alphabet $\Sigma$ is a function
  $G : \String \to \Set$.
\end{definition}

We say that a grammar $G$ associates to every string $w$ the set\footnote{We
  make little comment as to which foundations we use to encode sets. We argue\max{drop ``We argue''}
  that our construction is agnostic to the particular foundation, and so without
  loss of generality\max{incorrect use of ``without loss of generality''} we may take a class of small sets, a Grothendieck universe,
  a type theoretic universe, or any similar foundation. Here we choose a type
  theoretic universe.} $G w$ of parse
trees showing that $w$ matches $G$.

The simplest class of formal grammar on the Chomsky\max{first mention of Chomsky hierarchy, why not bring it up earlier when you are describing what a grammar is? Also cite}
hierarchy are the \emph{regular expressions}

\max{inaccurate reflection of common terminology. These are regular expressions,
  regular grammars usually means a CFG with a restriction on recursion that
  ensures you only allow Kleene star.}.

\begin{gather*}
  \mathsf{Char} \in \Sigma \\
  \mathsf{Reg} ::= \mathsf{Char}~|~I~|~\mathsf{Reg} \otimes \mathsf{Reg}~|~\mathsf{Reg} \oplus \mathsf{Reg}~|~\mathsf{Reg}^{*}
\end{gather*}

Here, we use a non-standard syntax inspired by linear logic. Disjunction of
regular expressions is written $\oplus$, and concatenation of regular
expressions is written like the linear tensor product $\otimes$. The
empty grammar is sometimes written in the same manner as the empty string,
$\varepsilon$, but we instead choose to represent the empty grammar by $I$. This
is also linear logic syntax to represent the unit $I$ of the operation $\otimes$.

To see an illustration of the notion of grammar given in \cref{def:grammar}, consider the regular expression
$G = (a \otimes b)^{*} \otimes a$ --- a sequence of 0 or more $ab$'s followed by
an $a$. We may think of $G$ as a function that takes in strings and outputs all
of the parse trees of that string. Passing the input string $baa$ through $G$
would output the empty set, because there are no $G$-parses of $baa$. However,
passing $aba$ through $G$ would output a singleton set, as there is only a
single $G$-parse of $aba$. Namely, the one that breaks up $aba$ as $(ab)a$.

Observe that for any input string $w$ for the above grammar $G$, $G(w)$ is
either empty or a singleton set. Either the string isn't parsed by $G$,
or it is parsed by $G$ for exactly one reason. That is, $G$ is
\emph{unambiguous}. If we instead consider the grammar
$H = (a \otimes b)^{*} \oplus (b \otimes a)^{*}$ there are now ambiguities. For
instance, what is the action of $H$ on the empty string $\varepsilon$?
The empty string $\varepsilon$ matches both $(a \otimes b)^{*}$ and $(b \otimes a)^{*}$, so
$\varepsilon$ can match either the left or the right pattern
of $H$. As a result, $H(\varepsilon)$ is a two-element set.

\[
  L_{G} = \{ w \in \String : G w \text{ is inhabited} \}
\]

As they are just sets, formal languages naturally are
endowed with a partial
order via the subset inclusion ordering. Similarly, formal grammars are
naturally endowed with the structure of a \emph{category} where a morphism
$f : G \to H$ is given by a \emph{family} of functions:

\[
  f^{w} : Gw \to Hw, w \in \String
\]

Intuitively, we read a morphism of grammars as a \emph{parse transformer}: $f$
translates a
$G$-parse of $w$ into an $H$-parse of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses. Isomorphisms in the category of
grammars are precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

The category of grammars is quite rich in structure. It can
equivalently be described as the functor category $\Set^{\String}$, where
$\String$ is a discrete category. That is, the objects in $\String$ are strings
and the only morphisms are identity morphisms. Such functor categories into
$\Set$ --- often called presheaf categories --- carry a remarkable amount of
structure. In fact, any presheaf category naturally models dependent type
theory, which we exploit to motivate the design of the formalism
presented in \cref{sec:tt}.

\section{A Type Theory to Reason About Formal Grammars}
\label{sec:tt}

\subsection{Syntax}
\label{subsec:syntax}

In this section we will walk through most of the syntax of our type theory. We
include the full syntax in \cref{sec:infer}.

The linear typing judgment in our syntax takes on the following schematic form,
\[
  \Gamma ; \Delta \vdash M : A
\]

Let us briefly digest what this judgment is trying to say. First, $A$ represents
a \emph{linear type} in our syntax. The intended semantics of these linear types
are formal grammars. That is, the linear typing system is designed to syntactically
reflect the behavior of formal grammars. For this reason, we may often
interchangeably use the terms ``linear type'' and ``grammar''. To see the
correspondence between types and grammars, note that the constructors for linear types
mirror operations on formal grammars. For example, we have linear type constructors
$\otimes$, $\oplus$, and $\amp$ which correspond to grammar concatenation,
disjunction, and conjunction, respectively. The other type constructors also have
corresponding grammar intuition which may be seen in \cref{subsubsec:other}.

$M$ is an inhabitant of type $A$, which we want to think of as a parse tree of
the grammar $A$. The core idea of this entire paper follows precisely from this
single correspondence: grammars are types, and the inhabitants of these types
are parse trees for the grammars!

$\Gamma$ represents a \emph{non-linear context} --- also referred to as an
\emph{intuitionistic} context. The non-linear fragment of the theory
corresponds to usual Martin-L\"of type theory. That is, the intuitionistic
contexts obey all of the usual structural rules --- contraction, weakening, and
exchange --- so we may use their contents as many times as we'd like.

On the other hand $\Delta$ represents a \emph{linear context}. These linear
contexts
behave substructurally --- in particular they have no contraction, weakening, or exchange
rule. This means that everything within $\Delta$ must be used \emph{exactly
  once} and \emph{in order of occurrence}. Thus, we can think of the linear
contexts as an ordered list of limited
resources. Once a resource is consumed, we cannot make reference to it again.
We may then
think of the variables in a linear context akin to building blocks for
constructing patterns over strings.

In the realm of parsing, our substructural
restrictions make it so that the linear contexts behave quite similarly to
formal grammars. For instance, linear context extension mirrors concatenation of
grammars which may be seen via the introduction rule for $\otimes$,

\[
  \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta , \Delta' \vdash e \otimes e' : A \otimes B}
\]

Just as two grammars may be concatenated, if we can build an $A$ with the
contents of $\Delta$, and a $B$ with the contents of $\Delta'$, then we we may
concatenate $\Delta$ and $\Delta'$ to build an inhabitant $A \otimes B$.

This is perhaps better illustrated through a concrete example. Consider the
regular expression $A = (a \oplus b) \otimes (b \oplus c)^{*}$ and the string
$w = abc$. We then have the following derivation that $w$ matches $A$,

\begin{equation}
  \label{eq:samplederiv}
  \inferrule
  {
    w_{1} : a \vdash \mathsf{inl}(w_{1}) : a \oplus b \\
    \inferrule
    {
      w_{2} : b \vdash \mathsf{inl}(w_{2}) : b \oplus c \\
      w_{3} : c \vdash \mathsf{inr}(w_{3}) : b \oplus c
    }
    {
      w_{2} : b, w_{3} :c \vdash \mathsf{cons}(w_{2},w_{3}) : (b \oplus c)^{*}
    }
  }
  {w_{1} : a, w_{2} : b, w_{3} : c \vdash \mathsf{inl}(w_{1}) \otimes \mathsf{cons}(w_{2}, w_{3}) : (a \oplus b) \otimes (b \oplus c)^{*} }
\end{equation}

When parsing $w$ into $A$ by hand, we perform the same process as reflected in
the derivation \cref{eq:samplederiv}. That is, we split the string $w$ up into pieces, one of which matches
$a \oplus b$, and and another that matches $(b \oplus c)^{*}$. When further
parsing this Kleene star bit, we break it up into smaller pieces each of which
matches the patterns $b \oplus c$ and then we combine them all into one parse of
the Kleene star $(b \oplus c)^{*}$. Note that this exact process is echoed in
the above proof tree.

\subsubsection{Dependence of the Linear on the Non-Linear}

We may now however ask, ``what are the $\Gamma$'s doing in the $\otimes$ introduction rule?''
Each linear context $\Delta$ is dependent on a non-linear context $\Gamma$, so
we may use variables drawn from $\Gamma$ when defining the types in $\Delta$.

In the grammar setting, this dependence lets
us use variables from regular dependent type theory to
keep track of data without \emph{consuming any resources}, so to speak. That is,
using a piece of a linear context consumes it, so we cannot linearly encode anything that
needs to be referenced more than once.

We will see a concrete instance of this dependence in \cref{subsubsec:pda}. In
that case, we encode the traces of a pushdown automaton as a grammar. When doing so,
we make use of our non-linear types to define a type of stacks. The fact that
our encoding of stacks is non-linear reflects
that a pushdown automaton may consume a character from the input string without
affecting the state of the stack.

\subsubsection{Other Type Constructors}
\label{subsubsec:other}
